--- 
title: "Biodose Tools"
author: "Alfredo Hernández"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: scrartcl
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: biodosimetry-uab/documentation
description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
apple-touch-icon: "favicon-highres.png"
apple-touch-icon-size: 120
favicon: "favicon.png"
---

# About


```{r echo=FALSE}
knitr::include_graphics("assets/home.svg", auto_pdf = getOption("knitr.graphics.auto_pdf", TRUE))
```

This project in an app to be used by biological dosimetry laboratories. Biodose Tools is an open-source project that aims to be a tool to perform all different tests and calculations needed. The app is developed with [R](https://www.r-project.org/about.html) [@R-base] together with [Shiny](https://shiny.rstudio.com) to offer an on-line, easy-to-use solution. Although the intention is to provide the application as a website, all R routines can be downloaded for improvement or personal use.

We also aim to clarify and explain the tests used and to propose those considered most appropriate. Each laboratory in its routine work should choose the optimum method, but the project aims to reach a consensus that will help us in case of mutual assistance or intercomparisons.

The project is initially developed by [RENEB](http://www.reneb.net) association, but contributions are always welcome.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
