## Software implementations {#intro-software}

```{r echo=FALSE, out.width = '75%', fig.align='center'}
knitr::include_graphics("images/intro/chronology.svg")
```

The tools to deal with some or all of the statistical procedures used in biological dosimetry have evolved since the 1980s [see @Ainsbury2009]. These tools ranged from adapted Microsoft Excel sheets to specific programs (the later summarised in \Cref{tab:software-summary}). The specific programs were based on closed source software and, for example, used a single methodology to calculate uncertainties. Thus, improvements were restricted to those who manage the source code, and the end-user cannot modify or implement any improvement. In addition, there is a possibility of misunderstanding or misuse of these closed programs.

```{r software-summary, echo=FALSE, purl=FALSE}
tibble::tribble(
  ~`Name`,         ~`Language(s)`,  ~`Framework(s)`, ~`Author(s)`,
  "MLPOL",         "Fortran 77",    "--",            "@Edwards1994",
  "MLREG",         "--",            "--",            "@BfS1996",
  "DOSGEN",        "Pascal",        "Turbo Pascal",  "@Garcia1996",
  "CABAS",         "Object Pascal", "Delphi",        "@Deperas2007",
  "Dose Estimate", "Object Pascal", "Delphi 6",      "@Ainsbury2010"
) %>%
  knitr::kable(
    caption = "List of historical analysis software developed specifically for radiation biodosimetry.",
    align = "l"
  )
```

Recently, using the R project for statistical computing [@R-base] some scripts have been written for biological dosimetry purposes. Although R programming is based on open source code, most end-users performing biodosimetry are usually not familiar with programming and mathematical or statistical modelling that is required to use or implement R source code. It is therefore required to provide tools with GUIs that are easy to use and that provide the required functionality to obtain reliable dose estimates.

In this paper we present {biodosetools} (styled as Biodose Tools), an R package developed using {shiny} [@R-shiny], to offer an online and easy-to-use solution to be used by biological dosimetry laboratories, as well as a tool for statisticians to manually perform the underlying calculations in R. The package is available from the Comprehensive R Archive Network (CRAN) at \url{https://CRAN.R-project.org/package=biodosetools} and GitHub at \url{https://github.com/biodosetools-team/biodosetools} and can be distributed under the GPL-3.0 license (GNU General Public License v3.0). Biodose Tools has been developed under the umbrella of RENEB (Running the European Network of Biological and retrospective Physical dosimetry).
