# Usage (R API) {#manual-use-r}

The following examples illustrate the functionality of {biodosetools}'s R API to perform dose-effect fitting and dose estimation for the dicentric and translocation assays. The equivalent examples using the {shiny} user interface can be found in Chapter \@ref(manual-use-shiny).

The full {biodosetools} R API reference detailing the available functions and parameters can be found on the project's website at https://biodosetools-team.github.io/biodosetools/reference/.

## Dicentrics dose-effect fitting

### Input count data

The first step is to input the count data. This step is accomplished in \proglang{R} by calling the `calculate_aberr_table()` function. This will calculate the total number of cells ($N$), total number of aberrations ($X$), as well as mean ($\bar{y}$), variance ($\sigma^{2}$), dispersion index ($\sigma^{2}/\bar{y}$), and $u$-value.
```{r dic-count-data}
count_data <- system.file("extdata", "count-data-barquinero-1995.csv",
  package = "biodosetools"
) |>
  utils::read.csv() |>
  calculate_aberr_table(type = "count")
```

```{r dic-count-data-r}
count_data
```

### Perform fitting

To perform the fitting in \proglang{R} we call the `fit()` function, whilst selecting the appropriate `aberr_module` (`"dicentrics"` in this example), `model_formula` (`"lin-quad"` or `"lin"` for LQ or L models, respectively), and `model_family` (`"automatic"`, `"poisson"`, or `"quasipoisson"`) parameters:
```{r dic-fit-results}
fit_results <- fit(
  count_data = count_data,
  model_formula = "lin-quad",
  model_family = "automatic",
  aberr_module = "dicentrics"
)
```

The `fit_results` object is a list that contains all necessary information about the count data as well as options selected when performing the fitting. This is a vital step to ensure traceability and reproducibility. Below we can see its elements:
```{r dic-fit-results-names}
names(fit_results)
```

In particular, we can see how `fit_coeffs` matches the results obtained in the UI (Figure \@ref(fig:sc-dic-fit-03)):

```{r dic-fit-coeffs-r}
fit_results$fit_coeffs
```

To visualise the dose-effect curve, we call the `plot_fit_dose_curve()` function:

```{r dic-fit-dose-curve, out.width="100%", fig.dpi=300, fig.align="center", fig.height=3, fig.width=6, fig.cap="Plot of dose-effect curve generated by {biodosetools}. The grey shading indicates the uncertainties associated with the calibration curve."}
plot_fit_dose_curve(
  fit_results,
  aberr_name = "Dicentrics"
)
```

## Dicentrics dose estimation

### Input case data

Next we can choose to either load the case data from a file or to input the data manually. The data is then parsed in \proglang{R} by calling the `calculate_aberr_table()` function. This will calculate the total number of cells ($N$), total number of aberrations ($X$), as well as mean ($\bar{y}$), standard error ($\sigma$), dispersion index ($\sigma^{2}/\bar{y}$), and $u$-value.
```{r dic-case-data}
case_data <- system.file("extdata", "cases-data-partial.csv",
  package = "biodosetools"
) |>
  utils::read.csv(header = TRUE) |>
  calculate_aberr_table(
    type = "case",
    assessment_u = 1,
    aberr_module = "dicentrics"
  )
```

```{r dic-case-data-r}
case_data
```

### Perform dose estimation

Finally, to perform the dose estimation in \proglang{R} we can call the adequate `estimate_*()` functions and parameters depending on the characteristics of the accident. In this example, we will use `estimate_whole_body_merkle()` and `estimate_partial_body_dolphin()`. First of all, however, we will need to load the fit coefficients and variance-covariance matrix from the previously calculated `fit_results`:
```{r dic-parse-fit-results}
fit_coeffs <- fit_results[["fit_coeffs"]]
fit_var_cov_mat <- fit_results[["fit_var_cov_mat"]]
```

After that is done, we can simply call `estimate_whole_body_merkle()` and `estimate_partial_body_dolphin()`:

```{r dic-dose-estimation-whole}
results_whole_merkle <- estimate_whole_body_merkle(
  case_data,
  fit_coeffs,
  fit_var_cov_mat,
  conf_int_yield = 0.83,
  conf_int_curve = 0.83,
  protracted_g_value = 1,
  aberr_module = "dicentrics"
)
```

```{r dic-dose-estimation-partial}
results_partial <- estimate_partial_body_dolphin(
  case_data,
  fit_coeffs,
  fit_var_cov_mat,
  conf_int = 0.95,
  gamma = 1 / 2.7,
  aberr_module = "dicentrics"
)
```

To visualise the estimated doses, we call the `plot_estimated_dose_curve()` function:

```{r dic-estimated-dose-curve, eval=FALSE, echo=TRUE}
plot_estimated_dose_curve(
  est_doses = list(
    whole = results_whole_merkle,
    partial = results_partial
  ),
  fit_coeffs,
  fit_var_cov_mat,
  protracted_g_value = 1,
  conf_int_curve = 0.95,
  aberr_name = "Dicentrics"
)
```

```{r dic-estimated-dose-curve-fix, echo=FALSE, message=FALSE, out.width="100%", fig.dpi=300, fig.align="center", fig.height=3, fig.width=6, fig.cap="Plot of estimated doses generated by {biodosetools}. The grey shading indicates the uncertainties associated with the calibration curve."}
plot_estimated_dose_curve(
  est_doses = list(
    whole = results_whole_merkle,
    partial = results_partial
  ),
  fit_coeffs,
  fit_var_cov_mat,
  protracted_g_value = 1,
  conf_int_curve = 0.95,
  aberr_name = "Dicentrics"
) |>
  fix_estimated_dose_curve_color_labels(
    breaks = c("Whole-body", "Partial-body"),
    labels = c("Whole-body (83\\%-83\\%)", "Partial-body (95\\%)")
  )
```


## Translocations dose-effect fitting

## Translocations dose estimation
