# Dicentric analysis {#stats-dicent}

## Dose-effect curve fitting

```{r include=FALSE}
library(dplyr)
```

Table \@ref(tab:count-data) gives example data used to construct dose-effect curves for low LET $\gamma$-radiation and high LET $\alpha$ radiation.

```{r count-data, echo=FALSE}
mytable <- data.table::fread(here::here("data/count-data-IAEA.csv"))

mytable <- mytable %>%
  as.data.frame() %>% 
  dplyr::mutate(
    N = 0,
    X = 0,
    DI = 0,
    u = 0
  ) %>%
  dplyr::select(D, N, X, dplyr::everything())

first_aberr_index <- 4
last_aberr_index <- ncol(mytable) - 2
num_rows <- nrow(mytable)

mytable <- mytable %>%
  dplyr::mutate(
    D = as.numeric(D),
    X = as.integer(X),
    N = as.integer(rowSums(.[first_aberr_index:last_aberr_index]))
  )

# Ugly method to calculate index of dispersion
for (row in 1:num_rows) {
  xf <- 0
  x2f <- 0
  # Calculate summatories
  for (k in seq(0, last_aberr_index - first_aberr_index, 1)) {
    xf <- xf + mytable[row, grep("C", names(mytable), value = T)][k + 1] * k
    x2f <- x2f + mytable[row, grep("C", names(mytable), value = T)][k + 1] * k^2
  }
  # Calculate variance and mean
  var <- (x2f - (xf^2) / mytable[row, "N"]) / (mytable[row, "N"] - 1)
  mean <- xf / mytable[row, "N"]
  # Save values into data frame
  mytable[row, "X"] <- as.integer(xf)
  mytable[row, "DI"] <- var / mean
  mytable[row, "u"] <- (var / mean - 1) * sqrt((mytable[row, "N"] - 1) / (2 * (1 - 1 / mytable[row, "X"])))
}

mytable %>% 
  knitr::kable(booktabs = TRUE, 
               caption = "[Download dicentrics distribution.](https://biodosetools-team.github.io/documentation/data/count-data-IAEA.csv)") #%>% 
  # kableExtra::kable_styling(latex_options = "basic")
```


For each dose analysed, total number of cells scored ($N$), total number of dicentrics observed ($X$), cell distribution of dicentrics and dispersion index ($\sigma^{2}/\bar{y}$) and $u$-test ($u$) are presented. $u$-values greater than 1.96 indicate overdispersion.

### Fitting method 
The technique suggested for determining the best fit coefficients is that of maximum likelihood [@Papworth1975; @Merkle1983]. Using this method, the best fit value for each coefficient is achieved by assuming a Poisson distribution and maximizing the likelihood of the observations by the method of iteratively reweighted least squares. For overdispersed (non-Poisson) distributions, as obtained after high LET radiation, the weights must take into account the overdispersion. If the data show a statistically significant trend of $\sigma^{2}/\bar{y}$ with dose, then that trend should be used. Otherwise, the Poisson weight on each data point should be divided by the average value of $\sigma^{2}/\bar{y}$.

```{r dic-coeffs, echo=FALSE}
mytable <- data.table::fread(here::here("data/dicentrics-coeffs.csv"))

mytable %>% 
  knitr::kable(booktabs = TRUE, 
               caption = "Dicentrics Coefficients")

```

### Goodness of fit 
<!-- The goodness of fit of the curve and significance of fitted $\alpha$ and $\beta$ coefficients should then be tested, for instance using the Chi-squared ($\chi^{2}$) test and an appropriate form of the $F$-test (e.g. $F$-test, $z$-test or $t$-test) respectively. Biodose Tools implements the $t$-test. -->

The goodness of fit of the curve and significance of fitted $\alpha$ and $\beta$ coefficients should then be tested, for instance using an appropriate form of the $F$-test (e.g. $F$-test, $z$-test or $t$-test) respectively. Biodose Tools implements the $t$-test.

Let $\hat{\theta}$ be an estimator of the parameter $\theta = \alpha, \beta, C$ in the fit model. Then the $t$-statistic for this parameter is defined as
\begin{equation} 
  t_{\hat{\theta}} = \frac{\hat{\theta}}{\widehat{se}(\hat{\theta})}
  (\#eq:t-test)
\end{equation}


<!-- If there is evidence of a lack of fit (i.e. the $\chi^{2}$ is greater than the degrees of freedom ($df$)), then the standard error should be increased by $(\chi^{2}/df)^{1/2}$. Many computer programs calculate SE values that are based on the sum of squares, instead of the Poisson estimate of the variance, which may lead to a false underestimation of the Poisson error. For this reason, when SE are calculated using this method, and the df is greater than the $\chi^{2}$, it is a good practice to increase the SE by $(df/\chi^{2})^{1/2}$. -->

The $p$-values of the $t$-test shown in Table \@ref(tab:dic-coeffs) indicate that the fitted data points were not statistically different from the observed ones confirming a good fit.

## Dose estimation
