# Dicentric analysis {#stats-dicent}

## Dose-effect curve fitting

Table \@ref(tab:count-data) gives example data used to construct dose-effect curves for low LET $\gamma$-radiation and high LET $\alpha$ radiation.

```{r count-data, echo=FALSE}
dic_count_data <- here::here("data/count-data-IAEA.csv") %>%
  utils::read.csv() %>%
  biodosetools::calculate_aberr_table(type = "count")

dic_count_data <- dic_count_data %>%
  as.matrix() %>%
  formatC(format = "f", digits = 3) %>%
  as.data.frame() %>%
  dplyr::mutate(
    dplyr::across(
      .cols = c("N", "X", dplyr::starts_with("C")),
      .fns = as.integer
    )
  ) %>%
  biodosetools:::fix_count_data_names(type = "count", output = "kable")

colnames(dic_count_data)[1] <- "$D \\text{ (Gy)}$"

dic_count_data %>%
  knitr::kable(
    booktabs = TRUE,
    caption = "Cytogenetic results obtained from blood samples irradiated with $\\gamma$-rays",
    # caption = "[Download dicentrics distribution.](https://biodosetools-team.github.io/documentation/data/count-data-IAEA.csv)"),
    align = "c",
    escape = FALSE
  )
```


For each dose analysed, total number of cells scored ($N$), total number of dicentrics observed ($X$), cell distribution of dicentrics and dispersion index ($\sigma^{2}/\bar{y}$) and $u$-test ($u$) are presented. $u$-values greater than 1.96 indicate overdispersion.

### Fitting method 
The technique suggested for determining the best fit coefficients is that of maximum likelihood [@Papworth1975; @Merkle1983]. Using this method, the best fit value for each coefficient is achieved by assuming a Poisson distribution and maximizing the likelihood of the observations by the method of iteratively reweighted least squares. For overdispersed (non-Poisson) distributions, as obtained after high LET radiation, the weights must take into account the overdispersion. If the data show a statistically significant trend of $\sigma^{2}/\bar{y}$ with dose, then that trend should be used. Otherwise, the Poisson weight on each data point should be divided by the average value of $\sigma^{2}/\bar{y}$.

```{r dic-coeffs, echo=FALSE}
mytable <- data.table::fread(here::here("data/dicentrics-coeffs.csv"))

mytable %>%
  knitr::kable(
    booktabs = TRUE,
    caption = "Dicentrics Coefficients",
    escape = FALSE
  )
```

### Goodness of fit 
<!-- The goodness of fit of the curve and significance of fitted $\alpha$ and $\beta$ coefficients should then be tested, for instance using the Chi-squared ($\chi^{2}$) test and an appropriate form of the $F$-test (e.g. $F$-test, $z$-test or $t$-test) respectively. Biodose Tools implements the $t$-test. -->

The goodness of fit of the curve and significance of fitted $\alpha$ and $\beta$ coefficients should then be tested, for instance using an appropriate form of the $F$-test (e.g. $F$-test, $z$-test or $t$-test) respectively. Biodose Tools implements the $t$-test.

Let $\hat{\theta}$ be an estimator of the parameter $\theta = \alpha, \beta, C$ in the fit model. Then the $t$-statistic for this parameter is defined as
\begin{equation} 
  t_{\hat{\theta}} = \frac{\hat{\theta}}{\widehat{se}(\hat{\theta})}
  (\#eq:t-test)
\end{equation}


<!-- If there is evidence of a lack of fit (i.e. the $\chi^{2}$ is greater than the degrees of freedom ($df$)), then the standard error should be increased by $(\chi^{2}/df)^{1/2}$. Many computer programs calculate SE values that are based on the sum of squares, instead of the Poisson estimate of the variance, which may lead to a false underestimation of the Poisson error. For this reason, when SE are calculated using this method, and the df is greater than the $\chi^{2}$, it is a good practice to increase the SE by $(df/\chi^{2})^{1/2}$. -->

The $p$-values of the $t$-test shown in Table \@ref(tab:dic-coeffs) indicate that the fitted data points were not statistically different from the observed ones confirming a good fit.

## Dose estimation
